A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2011-11-18 papers/contact/paper.tex
rewrite a bit of the contact paper Tested everything.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the use of the White Bear functional,
  based on Fundamental Measure Theory~\cite{roth2002whitebear} to
  predict contact densities for the inhomogeneous hard-sphere fluid.
\end{abstract}

@@

%% \begin{figure}
%% \includegraphics[width=5cm]{figs/free-energy}
%% \caption{A comparison of the various approximations to the free energy
%%   in the homogeneous limit.  Note that the White Bear functional uses
%%   a modified form of the Carnahan equation of state.  As expected,
%%   these are all identical, and we won't include this plot in the paper
%%   (but it's here to verify that our code is working).}
%% \label{fig:free-energy}
%% \end{figure}

The hard-sphere contact density---that is to say, the density of hard
spheres touching a given hard sphere---is important for the
association term in Statistical Associating Fluid Theory (SAFT).  For
a homogeneous hard-sphere fluid, the contact density---which is to
say, $n g_{HS}(\sigma)$, where $g_{HS}(\sigma)$ is the contact value
of the correlation function---is easily computed from the
Carnahan-Starling equation of state
\begin{align}
  g_{HS}(\sigma) &= \frac{1}{1-\eta} + \cdots
\end{align}
The contact density may be easily computed from a knowledge of the
contact theorem, which states that the pressure on any hard surface is
given by
\begin{align}
  p &= k_BT n_\textit{contact}
\end{align}
Since we are interested in the contact density at the hard-sphere
surface, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the free energy on hard sphere
radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
Since $\eta \equiv \frac{4\pi}{3} R^3$, we can work out that
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi (2R)^2$.  The area which is relevant
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
  &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  &= n k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
As it turns out, this is the standard answer for the contact value of
the autocorrelation function! $\ddot\smile$

@@

@@

\begin{widetext}
  \appendix

  \section{Partial derivatives of the White Bear free energy functional}

>>>>

%\newcommand{\derivation}[1]{#1} % Use this to show all derivations in detail
\newcommand{\derivation}[1]{} % Use this for nice pegagogical paper...

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the contact density of an inhomogeneous hard-sphere
  fluid, which is of particular interest, as it plays a critical role
  in Statistical Associating Fluid Theory (SAFT), which is the basis
  of a number of recent classical density functionals.  We derive a
  formula for the contact density from the White Bear version of the
  Fundamental Measure Theory functional~\cite{roth2002whitebear}, and
  test this functional agains Monte Carlo simulations.
  \textcolor{red}{Insert summary of results here...}
\end{abstract}

@@

The hard-sphere contact density---that is to say, the density of hard
spheres touching a given hard sphere---is important for the
association term in Statistical Associating Fluid Theory (SAFT).
For a homogeneous hard-sphere fluid, the contact density---which is to
say, $n g_{HS}(\sigma)$, where $g_{HS}(\sigma)$ is the contact value
of the correlation function---is well-known.  However, for
inhomogeneous fluids, the contact density has not been well studied.

\subsection{Association free energy of SAFT}
\newcommand\epsilonassoc{\ensuremath{\varepsilon_\textit{AB}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{AB}}}
\newcommand\ncontact{\ensuremath{n_\textit{contact}}}

The free energy term in Statistical Associating Fluid Teory (SAFT)
from which it derives its name is the association term, which accounts
for hydrogen bonding.  Hydrogen bonds are modeled as attractive
patches (``association sites'') on the surface of hard spheres.  These
sites represent protons or electron lone pairs, and have an attractive
energy $\epsilonassoc$ when two molecules are oriented such that the
proton of one overlaps with the lone pair of the other.  The volume
over which this interaction occurs is $\kappaassoc$, giving the
association term in the free energy has two empirical parameters fit
to experimental data.

The association free energy per unit volume has the form
\begin{align}
  f_\text{assoc} &= k_BT n\sum_A 
                  \left(\ln X^A - \frac{X^A}{2} + \frac12\right)
\end{align}
where the summation is over the association sites, and $X^A$ is the
fraction of association sites \emph{not} hydrogen-bonded.  The
fractions $X^A$ are determined by the self-consistent equations
\begin{align}
  X^A &= \frac{1}
  {1 + \sum_B n X^B \Delta^{AB}}\label{eq:X}
  \\
  \Delta^{AB} &= g(\sigma) \kappaassoc\left( e^{\epsilonassoc/kT} -
  1 \right)\label{eq:delta}
\end{align}
where $g(\sigma)$ is the correlation function evaluated at contact.
The product of density with correlation function evaluated at contact
gives the contact density, when we combine Equations~\ref{eq:X} and
\ref{eq:delta}:
\begin{align}
  X^A &= \frac{1}
  {1 + \sum_B \ncontact\kappaassoc X^B\left( e^{\epsilonassoc/kT} -
  1 \right)}
\end{align}
where the product $\ncontact\kappaassoc$ is the expected number of
molecules present in the association volume in the absence of the
association interaction.  In the SAFT-VR
model\cite{gil-villegas-1997-SAFT-VR}, this contact density is found
by adding a perturbative correction to the contact value for a hard
sphere fluid.

\subsection{Homogeneous limit}

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{3} R^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi (2R)^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

\section{Direct correlation function}

\textcolor{red}{FIXME WRITE THIS SECTION!} The direct correlation
function is readily accessible from density-functional theory, as the
second functional derivative of the free energy:
\textcolor{red}{(Something like...)}
\begin{align}
  -c^{(2)}(\xx_1, \xx_2) &= \frac{1}{k_BT}\frac{\delta^2 F_{ex}}{\delta n(\xx_1) \delta n(\xx_2)}
\end{align}
The direct correlation function is related to the ordinary correlation
function by the Ornstein-Zernike relation
\begin{align}
  h(\xx_1, \xx'_2) &= g(\xx_1, \xx_2) - 1 \\
  h(\xx_1, \xx_2) &= c(\xx_1, \xx_2)
    + \int n(\xx_3)c(\xx_1, \xx_3) h(\xx_2, \xx_3)d\xx''
\end{align}
From these two equation, we could in principle extract the correlation
function evaluated at contact from the DFT free energy functional.
The Ornstein-Zernike equation is usually solved by taking a Fourier
transform of all the relevant quantities.  In the inhomogeneous case,
however, this approach isn't workable, since the Fourier transforms
don't speed things up like you'd hope, and we don't have a periodic
system in general.

\textcolor{red}{It seems worth looking into this, to see what (if
  anything) we can do! I don't see a solution just now, but that
  doesn't mean there isn't one around somewhere.  In any case, just
  looking at the direct correlation function might give us some
  hints as to a different and interesting approximation for the
  contact density.}

@@

\appendix

\begin{figure}
\includegraphics[width=5cm]{figs/free-energy}
\caption{A comparison of the various approximations to the free energy
  in the homogeneous limit.  Note that the White Bear functional uses
  a modified form of the Carnahan equation of state.  As expected,
  these are all identical, and we won't include this plot in the paper
  (but it's here to verify that our code is working).}
\label{fig:free-energy}
\end{figure}

\begin{widetext}
  \section{Partial derivatives of the White Bear free energy functional}
